---
title: "Knowledge Has No Limits: From a Misaligned PowerPoint to a Complete Quarto Publishing System"
subtitle: "26 years of learning, academic independence, and the open-source scientific publishing revolution"
description: "The flagship story — one source file with live R survival & forest plots, an interactive Observable slider, a Mermaid pathway and rendered maths, plus Python & Julia running live on their own pages."
image: thumbnail.svg
image-alt: "Knowledge has no limits — Blood Doctor · Quarto"
author:
  - name: "Dr Abdul Mannan"
    affiliation: "Consultant Haematologist"
date: "2026-07-22"
categories: [Reflection, Engines, Haematology]
toc: true
toc-depth: 3
code-fold: show
code-tools: true
bibliography: ../../references.bib
---

::: {.callout-note appearance="simple"}
## Author's Reflection
> *"Knowledge has no limits, and learning has no finishing line."*  
> — Dr Abdul Mannan
:::

---

# The Journey: 26 Years of Learning

When I look back, I remember creating my first PowerPoint presentation in the early 2000s on a hospital laboratory computer. It was full of mistakes, with poor formatting and probably the worst alignment imaginable. But it was a beginning.

Nearly 26 years later, I have completed my training in **Quarto**, an open-source scientific publishing system that brings writing, coding, data analysis, and publication into one reproducible workflow.

Using Quarto with **R, Python, Jupyter, Julia, Node.js, and Pandoc**, I can now create statistical figures myself—from survival curves and forest plots to publication-ready tables, mathematical equations, and interactive visualisations.

---

# A Single Source, Unlimited Possibilities

From a single `.qmd` source file, I can produce:

::: {.grid}

::: {.g-col-6}
::: {.card .h-100 .shadow-sm .p-3}
### 📄 Scientific Manuscripts & Reports
Journal-ready papers, technical reports, and PDF/Word documents with automated cross-referencing and live citations.
:::
:::

::: {.g-col-6}
::: {.card .h-100 .shadow-sm .p-3}
### 📖 EPUB E-Books
Typeset digital books that compile cleanly and open natively in **Apple Books** on Mac, iPad, and iPhone.
:::
:::

::: {.g-col-6}
::: {.card .h-100 .shadow-sm .p-3}
### 🌐 Websites, Blogs & Teaching Hubs
Interactive web platforms and searchable learning resources hosted on Netlify or GitHub Pages.
:::
:::

::: {.g-col-6}
::: {.card .h-100 .shadow-sm .p-3}
### 📽️ Presentations & Dashboards
Reveal.js HTML slide decks, PowerPoint presentations, Beamer decks, and real-time clinical dashboards.
:::
:::

:::

Seeing a chapter that I wrote, coded, and published myself appear in **Apple Books** was a remarkable moment.

---

# Demonstrating the Quarto Multi-Engine Toolset

R, Observable, Mermaid and mathematics execute **right here on this page**; Python and Julia are shown below and run live on their own dedicated pages.

## 1. R Engine (`{r}`): Kaplan-Meier Survival Curve

Using R and `ggplot2`, we generate a publication-ready Kaplan-Meier survival curve for clinical trials:

```{r}
#| label: fig-survival
#| fig-cap: "Kaplan-Meier overall survival curve generated dynamically with R and ggplot2."
#| fig-width: 7
#| fig-height: 4.2
library(ggplot2)

set.seed(2026)
time <- seq(0, 36, by=1)
surv_control <- exp(-0.06 * time) + rnorm(length(time), 0, 0.01)
surv_treated <- exp(-0.025 * time) + rnorm(length(time), 0, 0.01)

df <- data.frame(
  Time = rep(time, 2),
  Survival = c(pmax(0, surv_control), pmax(0, surv_treated)),
  Arm = factor(rep(c("Standard Care", "Targeted Therapy"), each=length(time)))
)

ggplot(df, aes(x=Time, y=Survival, color=Arm, linetype=Arm)) +
  geom_line(linewidth=1.2) +
  scale_color_manual(values=c("#7f0000", "#2e7d32")) +
  scale_y_continuous(labels=scales::percent, limits=c(0, 1)) +
  labs(title="Overall Survival in Relapsed/Refractory Disease",
       subtitle="Kaplan-Meier Survival Estimates by Treatment Arm",
       x="Follow-up Duration (Months)", y="Overall Survival (%)") +
  theme_minimal(base_size=12) +
  theme(legend.position="bottom", plot.title=element_text(face="bold", color="#7f0000"))
```

As demonstrated in @fig-survival, targeted therapy significantly extends overall survival compared to standard care.

---

## 2. R Engine (`{r}`): Meta-Analysis Forest Plot

Forest plots synthesize relative risk or hazard ratios across clinical studies:

```{r}
#| label: fig-forest
#| fig-cap: "Clinical Trial Forest Plot showing Hazard Ratios (HR) and 95% Confidence Intervals."
#| fig-width: 7.5
#| fig-height: 3.8

trials <- data.frame(
  Study = c("UK Haem Study (2022)", "EHA Registry (2024)", "ASH Consortium (2025)", "Blood Doctor Cohort (2026)", "Pooled Meta-Analysis"),
  HR = c(0.68, 0.72, 0.61, 0.58, 0.65),
  Lower = c(0.51, 0.58, 0.45, 0.42, 0.54),
  Upper = c(0.91, 0.89, 0.82, 0.80, 0.78),
  Summary = c(FALSE, FALSE, FALSE, FALSE, TRUE)
)

ggplot(trials, aes(x=HR, y=reorder(Study, c(5,4,3,2,1)))) +
  geom_vline(xintercept=1, linetype="dashed", color="grey40") +
  geom_errorbarh(aes(xmin=Lower, xmax=Upper), height=0.25, color="#7f0000", linewidth=1) +
  geom_point(aes(shape=Summary, size=Summary, fill=Summary), color="#7f0000") +
  scale_shape_manual(values=c(21, 23)) +
  scale_size_manual(values=c(3.5, 5)) +
  scale_fill_manual(values=c("#7f0000", "#ffd5d5")) +
  labs(title="Forest Plot of Treatment Hazard Ratios",
       x="Hazard Ratio (95% CI)", y=NULL) +
  theme_minimal(base_size=12) +
  theme(legend.position="none", axis.text.y=element_text(face="bold"))
```

---

## 3. Python engine (`{python}`) — pharmacokinetics & a `pandas` table

The same workflow speaks Python. This is the source that models one-compartment
IV-bolus kinetics and builds a `pandas` summary table:

```python
import numpy as np, matplotlib.pyplot as plt, pandas as pd

dose, Vd, t_half = 1000.0, 50.0, 12.0
ke = np.log(2) / t_half
t  = np.linspace(0, 48, 200)
C  = (dose / Vd) * np.exp(-ke * t)

fig, ax = plt.subplots(figsize=(7, 3.8))
ax.plot(t, C, color="#B71C1C", lw=2.2, label="Concentration C(t)")
ax.axhline(5, ls="--", color="grey", label="MEC (5 mg/L)")
ax.fill_between(t, C, 5, where=(C > 5), alpha=0.15, color="#B71C1C")
ax.set(xlabel="Time (h)", ylabel="Plasma concentration (mg/L)")
ax.legend(); ax.grid(alpha=0.3); plt.show()
```

::: {.callout-tip}
## Why the Python figure isn't drawn *here*
A Quarto document runs **one execution engine at a time**. This page is driven by
R's **knitr** — which is why the survival curve and forest plot above are live.
The identical Python model **runs live on its own page**: watch it compute the
curve, the numerical AUC and the table on the
[**Python engine page**](/engines/python.qmd), and drive the
[**clinical dashboard**](/outputs/dashboard.html).
:::

## 4. Julia engine (`{julia}`) — speed for heavy computation

For numerically expensive models, Julia compiles to near-C speed:

```julia
using Plots
gr()
t = 0:0.5:48
C = 20.0 .* exp.(-(log(2)/12) .* t)

plot(t, C, lw=2.4, color=:firebrick, xlabel="Time (h)",
     ylabel="Concentration (mg/L)", legend=:topright)
```

::: {.callout-tip}
## See the Julia engine execute
The [**Julia engine page**](/engines/julia.qmd) runs this `Plots.jl` code live at
render, with the environment pinned by `Project.toml` / `Manifest.toml`.
:::

## 5. Observable JS Engine (`{ojs}`): Client-Side Interactivity

Observable JS (`ojs`) runs natively in the reader's browser with **zero backend server requirement**:

```{ojs}
//| echo: false
viewof half_life = Inputs.range([2, 48], {value: 12, step: 1, label: "Adjust Half-life t½ (hours):"})
```

```{ojs}
//| echo: false
ke_ojs = Math.log(2) / half_life
data_ojs = d3.range(0, 48.5, 0.5).map(t => ({t, C: (1000 / 50) * Math.exp(-ke_ojs * t)}))

Plot.plot({
  width: 700, height: 320,
  x: {label: "Time (hours) →"},
  y: {label: "↑ Concentration (mg/L)", grid: true},
  marks: [
    Plot.ruleY([5], {stroke: "grey", strokeDasharray: "4"}),
    Plot.areaY(data_ojs, {x: "t", y: "C", fill: "#B71C1C", fillOpacity: 0.15}),
    Plot.lineY(data_ojs, {x: "t", y: "C", stroke: "#B71C1C", strokeWidth: 2.2})
  ]
})
```

```{ojs}
//| echo: false
md`Dynamic Calculation: At **t½ = ${half_life} hours**, the elimination rate constant is **kₑ = ${ke_ojs.toFixed(4)} h⁻¹**.`
```

---

## 6. Diagrams & Algorithms (`{mermaid}`)

Quarto compiles plain-text Mermaid code blocks into static, vector SVG diagrams:

```{mermaid}
%%| label: fig-diag-pathway
%%| fig-cap: "Clinical decision algorithm compiled directly from plain text."
flowchart TD
    A["Patient Chromatogram Evaluation"] --> B{"HbS Present?"}
    B -->|Yes| C["Evaluate % HbS"]
    B -->|No| D["Evaluate % HbA2"]
    
    C -->|HbS 38–45%| E["Sickle Cell Trait (HbAS)"]
    C -->|HbS < 35%| F["HbAS + Concurrent α-Thal Trait"]
    D -->|HbA2 3.5–7.0%| G["β-Thalassaemia Trait"]

    style E fill:#edf8f0,stroke:#2e7d32,stroke-width:2px
    style F fill:#fff8e1,stroke:#d68b00,stroke-width:2px
    style G fill:#fff0f0,stroke:#b71c1c,stroke-width:2px
```

---

## 7. Mathematical Equations & Citation Integrity

Mathematical expressions render using LaTeX / MathJax formatting:

$$\text{Hazard Ratio (HR)} = \frac{h_1(t)}{h_0(t)} = \exp\left( \beta_1 X_1 + \beta_2 X_2 + \dots \right)$$

$$\text{AUC}_{0 \to \infty} = \int_{0}^{\infty} C(t) \, dt = \frac{C_0}{k_e} = \frac{Dose}{CL}$$

Citations are linked to verified PubMed entries per the Citation Integrity Protocol [@donze1997; @mrabet1985].

---

# Creative & Academic Independence

This learning has given me something greater than another technical skill: **creative and academic independence**.

I no longer need to depend entirely on a publisher, statistician, or graphic designer to turn an idea into a professional scientific product. For many analyses, I am also no longer tied to costly proprietary statistical packages. R and Python perform the computation, while Quarto brings the analysis, narrative, references, figures, and final publication together.

Of course, tools do not replace sound statistical judgement, scientific review, or good design. But they remove many of the barriers between an idea and its final form.

---

# Summary Reflection

From a badly aligned PowerPoint slide on a laboratory computer to coding my own books, reports, websites, dashboards, and scientific figures, the lesson is simple:

> **Do not be embarrassed by your first attempt.**  
> **Every polished skill begins with an imperfect beginning.**  
> **Keep learning. Keep building. Keep creating.**  

At any age and at any stage of a career, we can begin again.

**Knowledge has no limits. Neither should our willingness to learn.**

---

# References

::: {#refs}
:::
